---
title: "Zambia CSIP results"
author: "Michiel van Dijk, Amanda Palazzo and Petr Havlik"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  word_document:
    reference_docx: ../Support/word_styles_01.docx 
bibliography: ../Bib/CSIP_ZMB.bib
---

```{r setup, include=FALSE}
library(pacman)
library(readxl)
library(RColorBrewer)

p_load(rprojroot)
root <- find_root(is_rstudio_project)

source(file.path(root, "Scripts/support/get_dataPath.r"))

p_load(knitr)
knitr::opts_chunk$set(
  fig.width=8, fig.height=6,
  dpi = 300,
  echo=FALSE, warning=FALSE, message=FALSE,
  #fig.path = file.path(root,"Fig_tab_map/out-"),
  fig.path = file.path(projectPath, paste0("GLOBIOM/Results/", globiom_file, "/out-")))

p_load(pander)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)

source(file.path(root, "Scripts/report_input/results_bau.R"))
source(file.path(root, "Scripts/report_input/results_options.R"))
```

# Results

This section presents the results of our modelling effort. Quantifying the adoption of CSA practices with a global model, such as GLOBIOM, provides additional insights to the regional assumptions and offers context for the impact of global developments on the region. Additionally, a model represents a structured set of assumptions that can be used to challenge the ideas proposed by the visions and to check the visions for internal consistentency. The first part briefly describes the set of indicators we are going to use to assess future agricultural development of Zambia. The second part shows the model projections for the BAU scenario up to 2050 and compares those with the stakeholder vision. Finally, we illustrate the impact of adopting the selected CSA technologies on key agro-economic indicators.


## Indicators of impact

_Crop Yield._ Crop yield provides us with measure of the overall productivity of cropland. It is a key variable to assess and compare the impact of the selected CSA technologies. Increasing productivity is one of the pillars of CSA.

_Crop Production._ The supply of crop products, as an indicator does not enough provide context for achieving food security, as historically in many regions, increases in production have come from the expansion of cropland area rather than through intensification or yield improvements [@Alcott2005; @Byerlee2014; @Hertel2014]. 

_Livestock production and Livestock heads._ Livestock is an important component of the agricultural sector. It is modeled using a detailed representation of the global livestock sector where distinctions are made among dairy and other bovines, dairy and other sheep and goats, laying hens and broilers, and pigs [@Havlik2014; @Herrero2014]. Livestock production activities are defined by production systems [@Herrero2013]: for ruminants, grass based (arid, humid, and temperate/highlands), mixed crop-livestock (arid, humid, and temperate/highlands), and other; for monogastrics, smallholders and industrial. For each species, production system, and region, a set of input-output parameters is calculated based on the approach in @Herrero2013.

_Cropland and Land use change._  While productivity gains can increase profitability and may incentivize expanding cropland and grassland, in what is known as Jevon’s paradox [@Alcott2005; @Byerlee2014; @Hertel2014], we also examine the change in irrigated and rainfed cropland, grassland, forest, and other natural lands to determine to what extent productivity gains from irrigation is sparing land for protection of terrestrial ecosystems and climate stabilization. 

_Food Consumption._ One of the aims of CSA is to improve food security. We use kilocalorie availability per capita per day as an indicator to measure food security. This indicator considers the total amount of food products demanded in a region and translates the quantity into calories. 

_Trade._ Examining a region's imports or exports as a share of the domestic consumption allows us to assess the agricultural self-sufficiency or dependence on the outside world. 

_Agriculture, Forestry, and Other Land Use GHG emissions._ Examining emissions from the Agriculture, Forestry, and Other Land Use (AFOLU) sector allows us to contextualize climate stabilization through the impacts of expanded afforestation and biomass production for energy use, land use change such as deforestation and conversion of natural land to grassland and cropland, and the impacts of production of crop and livestock product. 


## Business as usual

The following section presents projections for a number of key agricultural variables up to the year 2050. The figures show the changes over time in terms of socio-economic and agricultural development for the coming three decades asssuming a Business as Usual (BAU) scenario for Zambia. As the impact of climate change is highly uncertain, which can have negative but also sometimes positive effects, the BAU represents a scenario of zero climate change. To show the impact of climate change, we added error bars to the figures, which show the range of projections under different climate change scenarios. As climate impact only starts to materialize in the long term, we only added error bars for the period 2020-2050. In the following section, we take a closer look at the impact of using various CSA technology options in comparison to the BAU scenario. This is followed by a section in which the results of a sensitivity analysis assessing the outcomes under different projections for population and economic growth. 

### Crop yields

GLOBIOM considers the technological improvement in crop productivity described in section 5 (Figure X) as a starting point for potential yield growth. Additionally, yields will reflect the change in cropland area between production systems and also the reallocation of crop production to more or less productive areas. Increases in production over time will come from a combination of the assumed yield improvements, expansion of the crop to current cropland including transitions between production systems, and expansion of the crop to new cropland taking into account the costs associate with land conversion and transitions between systems. Historically, for some crops an increase in production within Zambia has come from expansions in cropland area rather than through significant yield improvements [@Byerlee2014; @Hillocks2002]. In the second stakeholder workshop the modeling team recieved feedback that the initial projections of yield growth were unrealistic. We reexamined our assumptions and then adjusted them to be more in line with what the stakeholders envisioned.

Figure X depicts both the trend in historical crop yield from FAOSTAT and GLOBIOM projections for the four key food crops in Zambia: maize, cassave, groundnut and millit. Yield projections produced by GLOBIOM for the period 2000-2011 are very similar to the historical trend from FAOSTAT. Only in case of cassave, the yield projections might be too optimistic. In the BAU, maize yields increase by more than 65% from 2010 to 2050 reaching about 3.9 fm t/ha. This is comparable to results of @Kanyanga2013, who presents maize yield projections in the range of 3 fm t/ha for 2050. Cassava yields increase 45%, groundnut yields increase almost 50% and millet yields increase 61% from 2010 to 2050 (Figure X). The error bars illustrate the potential range in crop yield when running a range of different climate change scenarios and using a variety of crop models. In line with Figure X, the error bars in Figure X show that cassava yield is more likely to increase than decrease under climate change, while the reverse is true for the other three crops. In the most extreme scenario, maize yield is projected to be 3.3 fm t/y, which is 15% lower than the projected yield in the baseline of 3.9 fm t/ha. 

@CIAT2017 also present projections for the change in maize yield under an extreme climate scenario but it is not clear how they are calculated. The study reports a lower maize yield growth of 8,7pp in comparison to a no climate change scenario for the period 2020-2050. If we compare the growth rates between the BAU in the year 2020 (2.7 fm t/h) and the extreme climate scenario, a value of around 20pp is obtained, which is more than double the value reported in @CIAT. In other words, climate change scenarios considered in our analysis seem to cover a wider range (i.e. more extreme) shocks than the one considered by @CIAT. 

For none of the crops, the target of doubling productivity (in compared to average in the period 2009-2011) as stated in the vision is achieved. The gap is smallest for maize, followed by Cassava, while it is still very large for groundnuts and millet. 

_Figure X: BAU yield_

```{r fig_bau_yld}
fig_bau_yld
```

##### Source: FAOSTAT (historical) and IIASA (projections)


### Crop production 

The production of crops in Zambia is driven by changes in crop demand (also including foreign demand through trade) and supply side factors, in particular expected changes in yields. In 2000, maize accounted for nearly half of the total production of crop products in Zambia followed by sugar cane cassava, millet and groundnut. Under the BAU scenario, total crop production doubles by 2050 reaching around 6 Mt dm(Figure X). By 2050, maize is still the most produced crop in the region (production increases 191% in comparison to 2000). The GLOBIOM projections for 2000-2010 are in line with the historical trend, which shows gradual expansion in production of maize, sugar and several other crops. Only after 2010, production, in particular that of maize, started to increase rapidly, probably as a consequence of agricultural subsidy programs. As this implies a diversion of the BAU, which is unlikely to be maintained for the coming four decades, this fluctation is not taken into account by the model.

_Figure X: BAU crop production_

```{r fig_bau_crop_prod}
fig_bau_crop_prod
```

##### Source: FAOSTAT (historical) and IIASA (projections)
##### Note: Error bars indicate the range in total crop production as a consequence of climate change.


### Livestock production and Livestock heads

Under the BAU, the number of ruminant cattle increased with 22% from 2,7 million to 3.7 million heads for the period 2010-2050, which is still wel below the vision of 6 million heads in 2050 (Figure X). Measuring the development of the livestock sector in heads, however, presents only a partial picture. In reality, the increase in feed conversion efficiency over time will result in livestock units that produce more output over time (e.g. milk, eggs and meat). @Herrero2014 estimated that the feed conversion efficiencies in Sub-Saharan Africa would increase by 50% over the next 40 years under the middle of the road SSP2. The results in presented in Figure X, which shows an increase in ruminant meat production by more than 120% from 2010-2050. (Figure X). This comes from an increase in the feeding efficiencies of ruminant livestock.  

_Figure X: BAU Ruminant Meat Production_

```{r fig_bau_meat}
library(gridExtra)
grid.arrange(fig_bau_lvst, fig_bau_meat, ncol=2)
```

##### Source: IIASA


### Land use change

GLOBIOM endogenously models the global demand for land by considering the main users of land, cropland, grassland, short rotation plantation, managed forests and unmanaged forests and other natural land. Conversion from other natural and unmanaged forests will take place based on the economic profitability of the alternative land uses which include the costs to convert the land. An increase in the demand for food in Zambia is met by the region’s agricultural supply through productivity increases or though expansion of crop and grassland or by increasing the region's imports. 

In 2010, agricultural land accounts for less than 30% of the total land cover and cropland only 2%. 60% of the total land cover of Zambia in 2010 was forest area. Under the BAU, cropland area expands by a little more than 1 Mha by 2050, with total agricultural land (cropland and grassland) expanding by more than 3.2 Mha (Figure Xa). Most of the new agricultural land comes from forest (2.4 Mha), reducing the total forest area by 5%.

Figure Xb zooms in on the historical trend and BAU projection for cropland, showing area by crop as well as the vision target. According to vision, cropland should expand no more than 0.9 Mha thousand ha. FAOSTAT data indicates that the cropland area strongly increased from around 1.1 Mha in 2008 to almost 2 Mha in 2013, predominantly due to the expansion of maize area (and to a lesser extent cotton area). This pattern is a clear divergence from the BAU trend of the previous two decades. To account for this we use the average cropland for the period 2000-2013 as a baseline to calculate the vision target in 2050. The figure shows that under the BAU the total cropland will be still below the maximum expansion as stated in the vision. 

Climate change impact on land use change is relatively small. [@less land use because of cassava yield increase? CHECK]. Our maize area projections are more optimistic than @Kanyanga2013, who presents near static maize area projections up to 2050 of around 5 Mha. The main reason for the different finding is that we project an increase in maize exports and continues the present trend of positive nett exports (see below), while @Kanyanga2013 assume Zambia will become increasingly dependent on the imports of maize. We believe that our projected future is more in line with the trade pattern of Zambia in the last decade, which shows a positive nett export of maize.

_Figure X: Land use change_

```{r fig_bau_lc}
fig_bau_lc
```

##### Source: FAOSTAT (historical) and IIASA (projections)
##### Note: Historical data for forest and natural land are not available. Differences in land use change for all land use classes are relatively low and therefore not depicted in the figure. 


_Figure X: Cropland change_

```{r fig_bau_luc}
fig_bau_luc
```

##### Source: FAOSTAT (historical) and IIASA (projections)
##### Note: Total cropland excludes a number of crops (e.g. vegetables and fruits), which total land use is very small and cannot be modelled with GLOBIOM. Error bars indicate the range in total cropland as a consequence of climate change. 


### Food availability

According to the FAO, calorie availability in Zambia has hovered at or below 2000 kilocalories per capita per day since around 1985 and in 2010, the per capita calorie consumption was about 1900 [@FAO2017]. In the BAU scenario, calorie availability increases over the time by about @X due to the increase in the GDP per capita. By 2050, calories are about 50% higher than in 2010 (an increase of more than 930 kilocalories per day). This is well over the average caloric norm of 2,100 kcal/cap/day that is considered adequate to meet the energy needs for maintaining a healthy life and carrying out a light physical activity [@Haughton2009].In comparison to @Kanyanga2013, who projects kcal/cap/day values in the range of 1600-2400, our seem to be [@ADD]

_Figure X: BAU Calorie availability_

```{r fig_bau_cal}
fig_bau_cal
```

##### Source: FAOSTAT (historical) and IIASA (projections)


### Agricultural trade

Figure X shows the nett agricultural trade (exports minus imports) for a number of key agricultural commodities for 2010 and 2050 as well as the vision target of doubling (nett) exports in 2050. The results are mixed. For maize and cotton nett trade is positive in 2010 and an increase is expected for 2050. However, the increase will be lower than the target of doubling net trade. For cassava and sugarcane, who also exhibit a positive nett trade, a decrease is projeced. For millet, the positive next exports are even expected to become negative. Finally, for bovine meat, groundnuts, pig meat, rice, soya, sweet potatoes and wheat, net imports are expected to increase even further. The error bars indicate that climate change can have substantial impact on the trade balance of several crops, in particular cassava, maize, cotton, groundnuts and sugarcane. In the latter case, in the one extreme, nett trade is projected to be higher than the 2010 value, almost reaching the vision target, while in the other extreme, net trade will become negative.  


_Figure X: Nett agricultural trade_

```{r fig_bau_trade}
fig_bau_trade
```

### Agricultural and Land use, land-use change, and forestry (LULUCF) GHG emissions 

The greatest contributor to total GHG emissions in Zambia is the land use, land-use change, and forestry (LULUCF) sector, accounting for 86.6% of total emissions [@CIAT2017]. Agriculture and energy are responsible for 6% and and 6.5% of total emissions, respectively. Within LULUCF, 93% of the emissions are caused by the burning of biomass, while only 7% is due to the conversion of forest into cropland or any other type of land use coversion. GHG emissions in the agricultural sector or equally concentrated as the burning of savanna is responsible for 59% of total emissions, followed by emissions from enteric fermentation (12%), manure left on pasture (10%), cultivation of organic soils (10%), the use of synthetic fertilizers (5%) and other sectors (i.e. other use of manure, rice cultivation and crop residues).

Figure X and Figure X show the GHG emissions from the agriculture and LULUCF sector, respectively under a BAU scenario. GLOBIOM is only able to estimate the GHG emissions as a consequence of land use conversion and crop and livestock production. The burning of biomass and the burning of savanna, the largest contributors to GHG emissions, are not modelled. This implies that the figure only shows a very small share of total future GHG emissions in Zambia. For agriculture it is not possible to estimate emissions from the cultivation of organic soils and crop residue. Also due to differences in methodology it is not possible to directly compare the historical series for the LULUCF emissions with the GLOBIOM projections.   

In 2010, livestock accounted for more than 93% of the total GHG emissions from agriculture (not taking into account emssions from cultivatoon of organic soils and crop residue), while the remainder was caused by the crop sector, in particular the use of synthetic fertilizers (5.5%) and the production of rice (1.3%). In 2050 the share of emissions from synthetic fertilizer has increased to 14% of total emissions. The main cause for this is the shift toward more intensive production systems that make use of chemical fertilizers. The historical trend shows a much larger increase in total GHG emissions than our projections. A closer look at the data shows that according to FAOSTAT, the growth in emissions from synthetic fertilizers is much higher than the BAU projection. The reason for this is the recent and rapid growth in the application rate of fertilizers on maize, triggered by the national maize subsidy program.

LULUCF GHG emissions are projected to almost triple from around 2 MtCO2eq in 2010 to 5.5 MtCO2eq in 2050 caused by the conversion of forest to cropland and grassland. There is relatively little conversion of other natural land to cropland and grassland over the time period. 



_Figure X: BAU agricultural GHG Emissions_

```{r fig_bau_ag_emis}
fig_bau_ag_emis
```

##### Source: IIASA

_Figure X: BAU LULUCF GHG Emissions_

```{r fig_bau_lulucf_emis}
fig_bau_lulucf_emis
```

##### Source: IIASA


## CSA impact scenarios

This section reviews the impact of adopting the various CSA strategies discussed above on a number of key agricultural variables. In the figures below we present the BAU as "none" and then each bar represents a scenario in which one of the CSA with the adoption of the CSA practices. The error bars represent the maximum and minimum impacts from climate change as modeled by the biophysical crop models EPIC and and LPJmL. In the figures the adoption of agroforestry will appear as "af", conservation agriculture as "ca", residue retention as "rr", drought tolerant maize as "dtm" and expansion of irrigation as "ir".

Crop yields take into account the underlying assumption of the productivity of the CSA technologies as well as the assumed adoption rates of the technologies.In figure X we compare the average maize yield in 2050 under the different CSA scenario with the BAU average maize yield. For all of the CSA strategies except irrigation, the maize yields are higher, though residue retention has the largest yield impact. The error bar lines represent the impacts of the technologies under climate change. Residue retention, although the yields are highest, has the most sensitivity to climate change. 


### Crop yields

The biophysical impacts of climate change modeled by EPIC and LPJmL move in opposite directions. Maize yields from LPJmL are on average 21% lower than with no climate change assumptions. Maize yields from EPIC are on average about 1 percent higher than yields with no climate change assumptions. Drought tolerant maize performs better under climate change than the other CSA strategies, in part because of the design of the strategy where impacts from climate are dampened. Residue retention has slightly more negative impacts on maize yields, however, this could be due to the spillover productivity impacts from other crops which see less maize cultivated (which could lower the overall yield).  



_Figure X: Yield_

```{r fig_opt_yld}
fig_opt_yld_abs
fig_opt_yld_dif
```


### Crop production

maize production under the residue retention is 4% higher than the BAU, 2% higher under agroforestry and 1% higher under the conservation agriculture strategy. 

_Figure X: Production_

```{r fig_opt_prod}
fig_opt_prod_abs
fig_opt_prod_dif
```

##### Source: IIASA


### Land use change


In 2050, cropland area is lower than the BAU under the CSA strategy scenarios with agroforestry, conservation agriculture, residue retention and drought tolerant maize (Figure X). Irrigation infrastructure investments increase the cropland area expansion in 2050 by about 3%.  Residue retention saves the most land sparing appox. 64,000 ha of natural land in 2050 from conversion to cropland and grassland. Under BAU, irrigated area expands by 130% (reaching 120,000 ha of full control irrigation) and under a strategy where irrigation costs are partially subsidized, irrigated area expands 230% (an increase of an additional 52,000 ha). 

_Figure X: Land use change_

```{r fig_opt_land}
fig_opt_land_abs
fig_opt_land_dif
```

##### Source: IIASA


### Food availability

_Figure X: Food availability_

```{r fig_opt_calo}
fig_opt_calo_dif
```

##### Source: IIASA


### Trade

The increase in maize production from the adoption of the residue retention reduces the imports of maize dropping to only 5% of the domestic market share by 2050. The impacts on imports in the other CSA scenarios are limited since the relative increase in production compared to BAU is also about 1%. 

[Insert trade figure manually]

_Figure X: Trade_

```{r fig_opt_trade}
```

##### Source: IIASA


### GHG emissions

In figure x, we compare the difference in total emissions in 2050 under the BAU with the total emissions in 2050 under the different technologies. In almost all cases the adoption of the CSA technologies/practices result slightly less GHG emissions. Residue retention results in slightly more GHG emissions although this GHG accounting methodology does not account for the residue burning and therefore underestimates the potential GHG savings from residue retention. The impacts of GHG savings compared to the BAU are quite small because many of the proposed strategies were only applied to maize production (agroforestry, drought tolerant maize). Drought tolerant maize was also assumed to have no mitigation potential, however GHG mitigation is due to land sparing (less cropland area expansion) as a result of the more productive yielding maize. 

_Figure X: Emissions_

```{r fig_opt_emis}
fig_opt_emis_abs2
fig_opt_emis_dif
```

##### Source: IIASA


# Sensitivity analysis 

This section presents a sensitivity analysis to assess the uncertainty in socio-economic development and climate change on future development of the agricultural sector in Zambia. Instead of using the BAU scenario, which is based on SSP2, we now assume GDP, population and technical change are described by two opposing scenarios: SSP1, which assumes a sustainable and globalized world and SSP3, which assumes an unsustainable and fragmented world. The results for these two more extreme scenarios will provide insight into the 'bandwith' of plausible projections. For simplicity, we only present results for a selection of indicators, in particular crop production, crop yield and GHG emissions.



#######


[ SUMMARY table with which vision targer is met?]


# Discussion and limitations

In this paper we present a scenario analysis and model study to assess the impact of the adoption of CSA practices on agricultural development under climate change. When considering the results it is important to take the following limitations of the study into account. 

First, to model and compare the future impact of CSA practices, we need indicators on how crop yield, production and GHG emissions are affected with and without using these practices under equal conditions. Together with assumptions on adoption rates, change in key drivers and market effects, these coefficients will determine the final model projections. Such information can only be derived from detailed farm and plot level studies that use a (semi-) experimental setting to systematically compare the impact of different CSA options, while controlling for external conditions. Surprisingly, we found only a few studies that present this information and in most cases they covered only one CSA strategy and/or crop, were difficult to compare due to differences in methodologies and were conducted outside Zambia. Another problem is distinguishing between short and long-effects of most CSA practicies. This is important as it has been found that in some conditions the short-term productivity can even decrease under conservation agriculture [@Pittelkow2015]. Similarly, @Ajayi2009 show that it is important to use a five-year window to estimate the financial profitability of agro-forestry practices to account for the intertemporal aspects (e.g. fallow-crop cycles). At the moment we do not take such effects into account. Hence, the indicators we use are a proxy at best and therefore results should be interpreted with care. More farm and plot level studies to reveal the impact of CSA practices in a controlled setting are needed to improve the analysis.

Second, we are not able to fully take into account all spatially explicit factors that determine the impact of CSA strategies on agricultural production. As mentioned in the previous point, we have only very crude proxy indicators to differentiate the impact of CSA options in terms of crop yield and GHG emissions. In practice, the impact of adopting conservation agriculture, agro-forestry or drought tolerant maize varieties will be different across Zambia, depending among others on soil quality, climate and learning effects. Similarly, due to lack of high-resolution data on the location of small-farmers, spatial distribution of crops and costs of inputs, the model is run at a relatively coarse resolution (200 x 200 km), which limited spatial representation.

Third, the key source of GHG emissions in Zambia is the burning of biomass, which makes up around 80% of total GHG emissions in the country. The main cause for these emissions is the burning of forests for the production of charcoal, which is an important source of household energy supply in Zambia [@Dlamini2016]. Another important source of GHG emissions (3.5%) is the burning of savanna. At the moment GLOBIOM is not able to project the change in these two types of emission due to lack of information. For example, modelling the burning of biomass requires projections for the demand for charcoal, which, in turn, depends on the future supply of household energy in Zambia (e.g. hydropower, solar and wind energy). An interesting avenue for future research would be the design and modelling of such scenarios.

Finally, the crop models output we use to prepare our climate change crop yield shocks only capture the impact of anthro-pogenic climate change, i.e. the long-run changes in climate caused by human activity. This is different from strong variability in the climate (e.g. extreme weather events such as floods, droughts and hurricanes) that are also expected have considerable and negative impact on economic and agricultural development in Sub-Saharan Africa [@Ahmed2011; @Thurlow2012]. Modelling the effects of year-by-year variatons in climate requires the use of a stochastic modelling approach, which is currently under development.


# Conclusions

Under a business as usual future, the agriculture sector in Zambia grows significantly, which could mean doubling the agricultural output over the next 40 years. However, maize is currently and could remain the dominate crop in the region despite the vision for a diversified economy. The adoption of CSA strategies increase crop production but are limited to some extent by which crops can be targeted for CSA interventions and the realistic rate of adoption. 

With an improvement in the feed conversion efficiency of ruminant livestock production could increase significantly over the next 40 years with very little expansion of the cattle stocks. The development of the livestock sector will depend to some extent on the production of feed crops such as cassava and maize as well adopting the CSA strategy of retaining crop residues. 

Calorie availability improves in the region due an increase in per capita GDP growth and although the adoption of the CSA strategies have a limited impact on the calories consumption. Food demand in Zambia is also driven by the regional prices of crop products which depends the regional supply as well as the comparative advantage of supply of Zambia’s trading partners. Demand that cannot be met by production within Zambia will be imported, and if the region does not increase the productivity of its agriculture sector the share of imports may increase. CSA strategies that can increase the productivity and be adopted widely will help to reduce the share of food demand met by imports. 

GHG emissions will continue to rise in Zambia due to conversion of land for use by crop production and livestock grazing, which could be responsible for the deforestation of about 2% of Zambia’s forest reserves. CSA strategies that increase crop productivity reduce the land burden and may spare additional land from development. 

[ADD, recent increase in prod and fert use because of subsidied program? BAU may understimate prod and emis if the growth is increase. However, based on subsidies, question if they will be continued for decades. unlikely. ]


While the adoption of CSA strategies that have higher yields than traditional production methods will increase crop production, the types of that can be targeted by each strategy is important to consider especially in the context of a vision of Zambia that has a diversified economy.

Compared to conventional tillage, which often requires four to eight tractor passes in a typical growing season, no-tillage or zero tillage is significantly reducing the use of tractors and other heavy farm machinery and thus alleviating pollution through tractors and machinery. Furthermore, the increased levels of soil organic matter (SOM) and plant-available nitrogen typically found in CA soils greatly reduce the need for applying large amounts of chemical fertilizers, which causes pollution through N2 emission in the environment (Fig. 1.9). Thus, overall fossil fuel use and greenhouse gas emissions are greatly reduced under the CA (Niggli et al. 2009). https://link-springer-com.ezproxy.library.wur.nl/chapter/10.1007/978-981-10-2558-7_1



# References
