---
title: "Zambia CSIP"
author: "Michiel van Dijk"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
library(pacman)
p_load(rprojroot)
root <- find_root(is_rstudio_project)

p_load(knitr)
knitr::opts_chunk$set(
  fig.width=8, fig.height=6,
  dpi = 300,
  echo=FALSE, warning=FALSE, message=FALSE,
  fig.path = file.path(root,"Fig_tab_map/generated/out-"))

p_load(pander)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)

source(file.path(root, "Scripts/report_input/csa_options.R"))
```


# Note

CSA adoption scenarios for the period 2015-2050. 2015 values are based on RALS and supplied by Nick Sitko. 2050 values are based on outcomes of the CSIP 2nd scenario workshop. Diffusion curves are non-parametrically constructed using start and end year values combined with the general functional form of a generalised logistic function with a mid point of 2035 and a growth rate of between 0.3 and 0.4.

An annex is attached which includes the depicted values for each CSA practice for further use.

\newpage
_Conservation agriculture_

```{r}
fig_ca_proj
```

\newpage
_Minimum soil disturbance_

```{r}
fig_msd_proj
```

\newpage
_Crop rotation_

```{r}
fig_cr_proj
```

\newpage
_Residue retention_

```{r}
fig_rr_proj
```

\newpage
_Agro-forestry_

```{r}
fig_af_proj
```


\newpage
_Drought tolerant maize_

```{r}
fig_dtm_proj
```
