---
title: "Zambia CSIP results"
author: "Michiel van Dijk, Amanda Palazzo and Petr Havlik"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  word_document:
    reference_docx: ../Support/word_styles_01.docx
bibliography: ../Bib/CSIP_ZMB.bib
---

```{r setup, include=FALSE}
library(pacman)
library(readxl)
p_load(rprojroot)
root <- find_root(is_rstudio_project)

source(file.path(root, "Scripts/support/get_dataPath.r"))

p_load(knitr)
knitr::opts_chunk$set(
  fig.width=8, fig.height=6,
  dpi = 300,
  echo=FALSE, warning=FALSE, message=FALSE,
  #fig.path = file.path(root,"Fig_tab_map/out-"),
  fig.path = file.path(projectPath, paste0("GLOBIOM/Results/", globiom_file, "/out-")))

p_load(pander)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)

source(file.path(root, "Scripts/report_input/results_bau.R"))
source(file.path(root, "Scripts/report_input/results_options.R"))
```

# BAU scenario


In the following section, the business as usual future for Zambia, quantified using the GLOBIOM, is summarized to create a complete picture of the changes over time in terms of socio-economic and agricultural development. 




- Crop Production 

The production of crops in Zambia is driven by changes in crop demand and changes in crop prices in Zambia but to also from changes in the global crop demand. We can amining the relative differences in regional production allows us to identify regions where the scenario investments in irrigation infrastructure increase the region’s production contributions.

In 2010, maize accounted for nearly half of the total production of crop products in Zambia (46% of the total dry matter production), sugar cane accounted for 30% of the production, cassava accounted for 11%, and millet and groundnut together accounted for about 5% of the total production. 

In terms of cropland area, in 2010 more than half of cropland produced mazie, 15% produced cotton, 10% produced groundnut, 7% produced millet, and 4% produced cotton. Irrigated area (area under full-control irrigation) accounted for only about 4% of the total cropland area and primarily produced sugarcane and wheat. 

Under the BAU scenario, total crop production doubles by 2050 reaching nearly 6.6 Mt dm.  By 2050, maize is still the most produced crop in the region (production increases 221%), sugar cane increases 215% by 2050 and cassava production increases by 140% (Figure X).


_Figure X: BAU crop production_

```{r fig_bau_crop_prod}
fig_bau_crop_prod
```


- Crop Yields

GLOBIOM considers the technological improvement in crop productivity described in section 5.1 (Fig 2: Yield Projections) as a starting point for potential yield growth. Additionally, yields will reflect the change in cropland area between production systems and also the reallocation of crop production to more or less productive areas. Increases in production over time will come from a combination of the assumed yield improvements, expansion of the crop to current cropland including transitions between production systems, and expansion of the crop to new cropland taking into account the costs associate with land conversion and transitions between systems. Historically, for some crops an increase in production within Zambia has come from expansions in cropland area rather than through significant yield improvements [@Byerlee2014, @Hillocks2002].In the second stakeholder workshop the modeling team recieved feedback that the initial projections of yield growth were unrealistic. We reexamined our assumptions on the and adjusted the exogenous yield assumptions.

In the BAU, maize yields increase by more than 65% from 2010 to 2050 reaching about 3.3 fm t/ha. Cassava yields increase 45%, groundnut yields increase almost 50% and millet yields increase 61% from 2010 to 2050 (Figure X). 


_Figure X: BAU yield_

```{r fig_bau_yld}
fig_bau_yld
```




- Land use change

GLOBIOM endogenously models the global demand for land by considering the main users of land, cropland, grassand, short rotation plantation, managed forests and unmanaged forests and other natural land. Conversion from other natural and and unmanaged forests will take place based on the economic profitability of the alternative land uses which include the costs to convert the land. To expand cropland comes  An increases in the demand for food in Zambia is met by the region’s agricultural supply through productivity increases or though expansion of crop and grassland or by increasing the region's imports. 


In 2010, agricultural land accounts for less than 30% of the total land cover and cropland only 2%. 60% of the total land cover of Zambia in 2010 was forest area. Under the BAU, cropland area expands by a little more than 1 Mha by 2050, with total agricultural land (cropland and grassland) expanding by more than 3.2 Mha. Most of the new agricultural land comes from forest (2.4 Mha), reducing the total forest area by 5%. 



_Figure X: Cropland change_

```{r fig_bau_lc}
fig_bau_lc
```




- Livestock production and Livestock heads

Under the BAU, although the number of ruminant cattle increases only 22% from 2010 to 2050, the production of ruminant meat increases by more than 120% from 2010-2050 (Figure X). This comes from an increase in the feeding efficiencies of ruminant livestock. @Herrero2014 estimated that the feed conversion efficiencies in SSA would increase by 50% over the next 40 years under the middle of the road SSP2.  




_Figure X: BAU Ruminant Meat Production_
```{r fig_bau_meat}
fig_bau_meat
```

- Food Consumption

According to the FAO, calorie availability in Zambia has hovered at or below 2000 kilocalories per capita per day since around 1985 and in 2010, the per capita calorie consumption was about 1900 [@FAO2017]. In the BAU scenario, calorie availability increases over the time period by about due to the increase in the GDP per capita. By 2050, calories are about 50% higher than in 2010 (an increase of more than 930 kilocalories per day).



_Figure X: BAU Calorie availability_

```{r fig_bau_cal}
fig_bau_cal
```


- Trade
Although production of maize in Zambia increases, the demand for maize increases more significantly. In 2010, about imports represented about 9% of the market share of corn. By 2050, however, that share decreases to only 7%. 




- Agriculture, Forestry, and Other Land Use (AFOLU) GHG emissions

In 2010, livestock accounted for nearly 60% of the total GHG emissions from the AFOLU sector. GHG emissions over the period continue to grow and by 2050, the emissions from the conversion of forest to cropland and grassland become to the largest share of of emissions for Zambia. There is relatively little conversion of other natural land to cropland and grassland over the time period. N2O emissions from crop production also contiue to increase due to the expansion of cropland as well as the existing cropland shifting toward more intensive production systems that make use of chemical fertilizers. 



_Figure X: BAU Emissions_

```{r fig_bau_emis}
fig_bau_emis
```





# Comparison of technology options

Crop yields take into account the underlying assumption of the productivity of the CSA technologies as well as the assumed adoption rates of the technologies. 
In figure X we compare the average maize yield in 2050 under the different CSA scenario with the BAU average maize yield. For all of the CSA strategies except irrigation, the maize yields are higher, though residue retention has the largest yield impact. The error bar lines represent the impacts of the technologies under climate change. Residue retention, although the yields are highest, has the most sensitivity to climate change. 


_Figure X: Yield_

```{r fig_opt_yld}
#fig_opt_yld
fig_opt_yld_dif
#fig_opt_yld_abs
```

Corn production under the residue retention is 4% higher than the BAU, 2% higher under agroforestry and 1% higher under the conservation agriculture stratety. 

_Figure X: Production_

```{r fig_opt_prod}
#fig_opt_prod
fig_opt_prod_dif
```




_Figure X: Land_

In 2050, cropland area is lower than the BAU under the CSA strategy scenarios with agroforestry, conservation agriculture, residue retention and drought tolerent maize (Figure X). Irrigation infrastructure investements increase the cropland area expansion in 2050 by about 3%.  Residue retention saves the most land sparing appox. 64,000 ha of natural land in 2050 from conversion to cropland and grassland. 

```{r fig_opt_land}
#fig_opt_land
fig_opt_land_dif
```


The increase in corn production from the adoption of the CSA practices, further reduces the imports of maize which drops to only 5% of the domestic market share by 2050. 

_Figure X: Trade_
```{r fig_opt_trade}
fig_opt_trade
fig_opt_trade_dif
```





In figure x, we compare the difference in total emissions in 2050 under the BAU with the total emissions in 2050 under the different technologies. In almost all cases the adoption of the CSA technologies/practices result slightly less GHG emissions. Residue retention results in slightly more GHG emissions although this GHG accounting methodology does not account for the residue burning and therefore underestimates the potential GHG savings from residue retention. The impacts of GHG savings compared to the BAU are quite small because many of the proposed strategies were only applied to maize production (agroforestry, drought tolerant maize). Drought tolerant maize was also assumed to have no mitigation potential, however GHG mitigation is due to land sparing (less cropland area expansion) as a result of the more productive yielding maize. 




_Figure X: Emissions_

```{r fig_opt_emis}
#fig_opt_emis
fig_opt_emis_dif
```




#Sensitivity Analaysis
- Food Consumption

In the BAU scenario, calorie availability increases from 2010 to 2050 by almost 50%, but under SSP1 which has a higher growth in the GDP per capitacalories increase nearly 60% over the period (an increase of nearly 1000 kilocalories per day). Calorie availability in SSP3 only increases by 511 calories (27% increase from 2010). 



```{r fig_opt_cons}
#fig_opt_cons
#fig_opt_cons_dif
```


