---
title: "Zambia CSIP results"
author: "Michiel van Dijk, Amanda Palazzo and Petr Havlik"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  word_document:
    reference_docx: ../Support/word_styles_01.docx
bibliography: ../Bib/CSIP_ZMB.bib
---

```{r setup, include=FALSE}
library(pacman)
library(readxl)
p_load(rprojroot)
root <- find_root(is_rstudio_project)

source(file.path(root, "Scripts/support/get_dataPath.r"))

p_load(knitr)
knitr::opts_chunk$set(
  fig.width=8, fig.height=6,
  dpi = 300,
  echo=FALSE, warning=FALSE, message=FALSE,
  #fig.path = file.path(root,"Fig_tab_map/out-"),
  fig.path = file.path(projectPath, paste0("GLOBIOM/Results/", globiom_file, "/out-")))

p_load(pander)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)

source(file.path(root, "Scripts/report_input/results_bau.R"))
source(file.path(root, "Scripts/report_input/results_options.R"))
```

# BAU scenario


In the following section, the business as usual future for Zambia, quantified using the GLOBIOM, is summarized to create a complete picture of the changes over time in terms of socio-economic and agricultural development. 




- Crop Production 

The production of crops in Zambia is driven by changes in crop demand and changes in crop prices in Zambia but to also from changes in the global crop demand. We can amining the relative differences in regional production allows us to identify regions where the scenario investments in irrigation infrastructure increase the region’s production contributions.

In 2010, corn accounts for 



_Figure X: BAU crop production_

```{r fig_bau_crop_prod}
fig_bau_crop_prod
```



- Crop Yields




_Figure X: BAU yield_

```{r fig_bau_yld}
fig_bau_yld
```




- Cropland and Land use change



- Livestock production and Livestock heads



_Figure X: BAU lvst_

```{r fig_bau_lvst}
fig_bau_lvst
```




- Agriculture, Forestry, and Other Land Use (AFOLU) GHG emissions

_Figure X: BAU Emissions_

```{r fig_bau_emis}
fig_bau_emis
```



- Food Consumption

According to the FAO, calorie availability in Zambia has hovered at or below 2000 kilocalories per capita per day since around 1985 and in 2010, the per capita calorie consumption was about 1900 [@FAO2017]. In the BAU scenario, calorie availability increases over the time period by about due to the increase in the GDP per capita. By 2050, calories are about 50% higher than in 2010 (an increase of more than 930 kilocalories per day).



_Figure X: BAU Calorie availability_

```{r fig_bau_cal}
fig_bau_cal
```


- Trade



# Comparison of technology options

_Figure X: Yield_

```{r fig_opt_yld}
fig_opt_yld
fig_opt_yld_dif
fig_opt_yld_abs
```


_Figure X: Emissions_

```{r fig_opt_emis}
fig_opt_emis
fig_opt_emis_dif
```


_Figure X: Land_

```{r fig_opt_land}
fig_opt_land
fig_opt_land_dif
```


_Figure X: Price_

```{r fig_opt_price}
fig_opt_price
fig_opt_price_dif
```


_Figure X: Trade_

```{r fig_opt_trade}
fig_opt_trade
fig_opt_trade_dif
```

_Figure X: Livestock_

```{r fig_opt_lvst}
fig_opt_lvst1
fig_opt_lvst_dif1
fig_opt_lvst2
fig_opt_lvst_dif2
```


- Food Consumption

In the BAU scenario, calorie availability increases from 2010 to 2050 by almost 50%, but under SSP1 which has a higher growth in the GDP per capita (@add) calories increase nearly 60% over the period (an increase of nearly 1000 kilocalories per day). Calorie availability in SSP3 only increases by 511 calories (27% increase from 2010). 




_Figure X: Consumption_

```{r fig_opt_cons}
fig_opt_cons
fig_opt_cons_dif
```



_Figure X: Production_

```{r fig_opt_prod}
fig_opt_prod
fig_opt_prod_dif
```