---
title: "A quantitative impact assessment of Climate Smart Agricultural strategies on agricultural development in Zambia"
author: "Michiel van Dijk (IIASA, vandijkm@iiasa.ac.at), Amanda Palazzo (IIASA), Stefan Frank (IIASA), Tamas Krisztin (IIASA) and Petr Havlik (IIASA)"
date: "`r format(Sys.time(), '%B %d, %Y')`"
abstract: "Abstract: Climate change will have substantial and negative consequences for agricultural development in Zambia. The adoption of climate smart agriculture (CSA) is widely regarded as an approach to increase resilience of smallholders against climate change, while at the same time contributing to food security and climate mitigation. In this study we assess and compare the impact of adopting seven CSA options on future agricultural development in Zambia. The analysis is conducted using a partial-equilibrium agricultural sector model (GLOBIOM) that can assess future agricultural development and land use change under socioeconomic development and climate change impacts. To reflect local conditions, crop yield, production and land use information in GLOBIOM is updated with recent agricultural statistics from Zambia. The impact of adopting CSA options on agricultural development is contrasted with a business as usual scenario that assumes only limited expansion of CSA practices and a 'vision' of agricultural development based on policy documents and stakeholder input. The results of this study can be used to inform strategic and investment decisions on the support of CSA in Zambia." 
output: 
  word_document:
    reference_docx: ../../Support/word_styles_01.docx
bibliography: ../../Bib/CSIP_ZMB.bib
---

```{r setup, include=FALSE, cache=TRUE}
library(pacman)
library(RColorBrewer)

p_load(rprojroot)
root <- find_root(is_rstudio_project)

source(file.path(root, "Scripts/support/get_dataPath.r"))

p_load(knitr)
knitr::opts_chunk$set(
  fig.width=8, fig.height=6,
  dpi = 300,
  echo=FALSE, warning=FALSE, message=FALSE,
  fig.path = file.path(root,"Fig_tab_map/generated/out-"))

#  fig.path = file.path(projectPath, paste0("GLOBIOM/Results/", globiom_file, "/out-")))

p_load(pander)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)

source(file.path(root, "Scripts/paper_input/results_bau.R"))
source(file.path(root, "Scripts/paper_input/yecc.R"))
source(file.path(root, "Scripts/report_input/results_options.R"))

```


# Introduction

- Refer to de pinto study and fact that there are only very few model studies on CSA impact
- Use CCAFS working paper on climate risk Zambia
- Change diffusion curves so that adoptation in say 2100 becomes 100% and the steep part of the curve cuts through the adoption rate in 2050. Saturation level is 100% not adoption rate we set for 2050
- Apply simple +X tons CO2 carbon seq for agro-forestry and perhaps also other crops => use coefficients from FAO model? = see excel file. 


The government of Zambia highlighted the linkages and interdependencies between climate change and agricultural development in most recent development and policy plans. As the majority of people are living in rural areas and land use change contribute the largest part to greenhouse gas emissions, is essential to develop an effective and coherent approach to deal with the impact of climate change on the agricultural production and vice versa. 

At present, the agriculture sector development strategies do not provide a coherent future roadmap to deal with climate risk. To inform a structured approach, the government of Zambia, with support of the World Bank, is in the process of drafting the first Zambia Climate Smart Investment Plan (CSIP). The aim of this study is to contribute to the design of the CSIP by means of quantitative scenario modelling exercise using the GLOBIOM model developed by the International Institute for Applied Systems Analysis (IIASA). GLOBIOM is a spatially-explicit partial-equilibrium agricultural sector model that can assess future agricultural development and land use change under climate change impacts. For this study the model has been modified and updated with country specific input information in order to better represent the agricultural sector in Zambia. The quantitative scenario analysis supports the identification and prioritization of key policy interventions and investments to inform agricultural sector and climate change policy planning in Zambia for the medium to long term. 

The GLOBIOM model has been modified and updated to simulate a selective number of scenarios that represent possible future agricultural pathways. The scenarios are designed to capture both uncertainties in the drivers of future agricultural development (e.g. population growth, economic development, technical change and climate change) as well as a number of policy options (e.g. adoption of climate smart technologies and practices). The scenarios are contrasted with desired agricultural targets (referred to as the vision) that are taken from policy documents and stakeholder input. To address the substantial uncertainties related to climate change impact and socioeconomic drivers, a sensitivity analysis is conducted in which yield, GDP and population projections are set to plausible upper and lower boundaries. 

The formulation of the scenarios builds upon the ideas and views that were expressed by a wide group of stakeholders from various Ministries, NGOs, policy think tanks and international organizations at two workshops. At the first workshop that took place in Lusaka on the 16th and 17th of October, the main drivers of agricultural development and land use change were discussed and different policy options and strategies were proposed to address climate smart agriculture in Zambia. Stakeholders also prepared a vision statement for climate smart agriculture in 2050. The output of the stakeholder workshop as well as a selection of recent policy sector and planning documents were used to design the scenarios. At the second workshop that took place in Lusaka on the 17th of April, 2018, preliminary results were presented, discussed and validated. This document presents the final scenarios and results that incorporate comments by stakeholders.


# Literature review

A few papers assess the impact of climate change on agriculture in Zambia. @Jain2007 focuses on the economic impact of climate change using the Ricardian method. In this method, a regression model is estimated with net farm revenue as independent variables and a set of climate, hydrological, soil and socioeconomic variables as explanatory variables. The analysis is conducted on a sample of 1015 households, including small-, medium- and large-scale households from 30 out of the 72 districts. The results point out that farm revenue is negatively affected by an increase in the November-December mean temperature and a decrease in the January-February mean rainfall, while a positive effect is found for an increase in the January-February mean temperature. This suggests that agriculture in Zambia will be affected by expected future change in rainfall and temperature induced by climate change. A major drawback of the Ricardian approach is that it fully relies on historical cross-section data and therefore is not able to assess the impact of future changes in the agricultural development related to economic growth and technical change.

@Thurlow2012 use a combination of a hydro-crop (HC) model and dynamic computable equilibrium (CGE) model to assess the impacts of future climate change on economic growth and household poverty in Zambia. The HC model is applied to simulate the impact of climate change on crop yield, whose results are fed into the CGE model to capture the economic effects. The HadCM3-B1a scenario, which represents a future climate where rainfall declines and temperatures rise, and two more extreme scenarios (2% increase in temperature combined with 15% rainfall above and below the historical trend) are used to model climate change up to 2025. The HC model indicates that in the HadCM3-B1a scenario  average maize yields will be reduced by around 1% in four out of the five agro-climatic zones in comparison to the historical trend. The exception is the Western part of Zambia, where maize yields are expected to increase slightly. Yield change is larger (in the range of -6% to +4%) in the more extreme climate scenarios. Incorporating the effects of climate change in the CGE model and assuming the continuation of historical climate variability, GDP is projected to decrease between -0.29%-point and -1.31%-point per year, while the poverty rate is expected to increase between 1.69%-point and 7.23%-point, all in comparison to a normal rainfall scenario. The simulation assumes that GDP growth will not increase between the baseline and 2025. The results are therefore net GDP growth as a consequence of structural change in the economy over time.

@Kanyanga2013 combine a crop model (DDSAT) and global economic model (IMPACT) to investigate the impact of climate change on future food security in Zambia. IMPACT is a partial equilibrium model that is used to simulate agricultural development at the global level (but including national representation) under a range of future scenarios. IMPACT models the supply and demand for 39 crop products at a food production unit (FPU) level (intersection of national boundaries and river basins). The spatial resolution for the modeling the impacts of climate change on the agriculture sector in Zambia is 1 FPU. For this study, three socioeconomic (pessimistic, baseline and optimistic) and four climate change scenarios (CNRM-CM3, CSIRO- mark3, ECHAM5 and MIROC3.2) were combined to account for the large uncertainty in both types of drivers. All three socioeconomic scenarios show an increase in population and GDP in 2050 in comparison to the base year. The crop growth simulations broadly indicate that maize yield gain of sometimes over 25% is expected to occur in Western province, the eastern half of North Western Province, Copperbelt Province, and most of Northern and Luapula Provinces. At the same time, yield losses, sometimes in excess of 25%, are expected in Southern Province and parts of Eastern Province as well as other places scattered throughout the country. The results from IMPACT show that in all scenarios malnutrition among children under five years of age is decreasing and the availability of kilocalories per capita is decreasing between 2010 and 2050 although change is less positive in the pessimistic socioeconomic scenario. The model also shows that production and yield of maize is increasing and  net exports of nearly all modeled crops are declining. All results vary to some extent, depending on the climate scenario.

@CIAT2017 analyses the effects  of  climate  change  on  agriculture  in Zambia over the period 2020-2050, also using the IMPACT model. The impact of climate change on three variables is assessed, namely net trade, crop area (or livestock numbers) and yields, for scenarios with and without climate change. SSP2 is used as a reference scenario for no climate change but it is not clear from the document which crop model, radiative forcing scenario and global circulation model combination has been used for the climate change scenario. The outcomes suggest the selection of an extreme climate change scenario. The model projects that net trade will decrease for most agricultural commodities as a consequence of climate change. Crop yield are  projected  to  grow  over  the
period  2020  to  2050  under  both  scenarios but growth rates are lower for a number of crops including: cassava (2.9pp), cotton (6.5pp), groundnut (5.2pp), maize (8.7pp) and soy bean (3.0pp) while rice yield is expected to increase under climate change (3.4pp). Climate change impacts on crop area are similarly mixed. Land areas  cultivated  with groundnut, maize, and rice are expected to be larger under climate change, while the areas under cassava, cotton and vegetables  are projected to  increase  under both scenarios but less when there is climate change.


# Methods

## Global-to-local modelling approach

Drivers of agricultural development in Zambia include both local and global factors. At the local level, national policies (e.g. agricultural support and investment plans), population growth, economic growth and climate change impact are key drivers, while at the global level international trade is an important factor. It is the interplay and interaction between these two sets of drivers that eventually will determine the future development trajectory of agriculture in Zambia. To capture all drivers, we extend a global land use model (GLOBIOM) with a detailed representation for Zambia. GLOBIOM [@Havlik2014] is a spatially explicit partial equilibrium model of the global forest and agricultural sectors. It has been used extensively to analyse the medium and long run effects of climate change [@Leclere2014; @Havlik2015a], deforestation [@Schmitz2014], food security [@Valin2014a; @Palazzo2017] and bio-fuel policies [@Valin2013b; @Havlik2011; @Mosnier2013; @Frank2013] at the global and regional level. The model has also been used for a number of in-depth country case-studies, including a deforestation assessment for Congo [@Mosnier2014a] and Brazil [@Soterroni2018], and and analysis of crop production potential for Russia and Ukraine [@Deppermann2018]. The supply side of the model is based on a bottom-up approach (from land cover, land use, and management systems to production and markets). The agricultural and forest productivity is modeled at the level of grid cells of 5x5 to 30x30 arc-minutes, input from crop and forest models. Demand is modeled at the regional level using 30 economic regions which consist of single countries or bundles of countries based on the the food balance sheets developed by the FAO [@FAO2017]. For the purpose of this study, Zambia was separated from Southern African region in GLOBIOM. To improve the representation of Zambia, agricultural production statistics, land cover maps and, data on irrigation, costs and livestock were added [@which one?]. The Zambia region in GLOBIOM is modeled and optimized at 32 land units (LUs) that consider differences in agro-ecological zones (AEZ) for a total of 57 LU/AEZ combinations. @Annex A provides additonal information on the GLOBIOM model. 

To model the impact of the adoption of different CSA practices on the productivity of the agriculture sector as well as the potential for CSA mitigation of GHG emissions, we [elaborate on how CSA strategies are modelled, referrting to ISLAM: Global FoodSecurity10(2016) 63–70 and Rosegrant diffusion of technology report, who uses similar approach of diffusion curves]. 

We also build on the work of @Frank2018 and @Beach2015, who assesses the impact of a range of climate change mitigation options on agricultural production and GHG emissions at the global level. To further develop the mitigation options for Zambia, we refined the assumptions on yield and mitigation potential for several mitigation technology options based on stakeholder workshops a literature review (See below).^[@Two stakeholder workshops were organised to review and discuss relevant CSA options for Zambia: (1) and (2) April 2018, Lusaka]

The emission savings for each technical mitigation option are multiplied by GLOBIOM production system specific (crop- or livestock sector) emission factors to calculate absolute emission savings. This modelling approach explores the impact of the adoption of the mitigation options and therefore does not consider the direct costs and labor costs, change in input costs of each mitigation technology.  


## Scenario design

###	Business As Usual and climate scenarios

To assess the impact of adopting CSA practices on key agricultural indicators, we start with designing a Business As Usual (BAU) scenario, which assumes that trends that are typical of recent decades will continue in the future. As such it presents the baseline to which CSA scenarios can be compared. To model a business as usual (BAU) scenario for Zambia, we use the SSP2: Middle of the Road scenario [@Fricko2017] from the the Shared socioeconomic Pathways (SSPs), which is regarded as the BAU scenario [@REF]. The SSPs have been developed as a backbone for climate change related assessments by a consortium of researchers [@Kriegler2012; @VanVuuren2017a; @ONeill2017a] and are used in nearly all recent climate change impact assessments [@REF]. 

The SSPs include projections for key drivers, including population and income growth, which are available at country level, including Zambia. The SSPs do not include projections for crop yields, which are essential to model future agricultural productivity growth. We use yield projections from @Herrero2014 that are estimated using the historical relationship between GDP growth and crop yield increase. We assume that these projections represent intrinsic productivity rates and reflect the increase in yield as a consequence of advances in knowledge and new technologies. They do not incorporate the yield shocks that occur as a consequence of climate change nor the additional yield that is associated with the adoption of (some) CSA practices. Finally, to model the BAU scenario, we also make assumptions on area of protected land, change in diets, food loss and waste, and trade at the global level [see @Popp2017 for details]. Annex X depicts the SSP2 projections for population, income and yield growth used for the modelling for Zamia and the rest of the world.

The SSPs can be combined with assumptions on climate outcomes, the so-called representative concentration pathways (RCPs), to derive a matrix that reflects an elaborate scenario framework to assess the impact of climate change and its mitigation under a variety of socioeconomic conditions [@VanVuuren2014]. To capture the uncertainty related to climate change, we use climate change projections (i.e. temperature and precipitation change) from the of the the Inter-Sectoral Impact Model Intercomparison Project [@Warszawski2014], which combines the results from five General Circulation Models (Figure @S1a-e). To assess the impact of climate change on agricultural production, the GCM results are combined with crop models that project the change in crop yield as a consequence of climate change. We only use the climate change projections for RCP 8.5, which is the most extreme climate scenario [@Riahi2011].To address the uncertainty in crop model design and parameterization, we use crop yield projections from two crop models: EPIC and LPJmL. 

Figure @Add shows the average (2 crop models x 5 GCMs) simulated crop yield change for a selection of crops in Zambia for the period 2010-2050. The impact of climate change on yield also depends on the farming system (GLOBIOM distinguishes between irrigated, high input, low input and subsistence rainfed farming). Irrigated and high input systems are less vulnerable to climate change because water and nutrients (i.e. fertilizer) are not a limiting factors. As the majority of farmers in Zambia are subsistence farmers we only present the related crop yield results. The figure shows the large uncertainty associated with the crop yield simulations, which are both caused by the large differences in climate change scenarios as well as differences in crop model projections [also see @Rosenzweig2014]. The uncertainty is particularly large for beans, cassava and potatoes. Out of the 11 crops, climate change is expected to result in a negative yield shock for six crops as illustrated by a median value below zero. For five crops (barley, beans, cassava, rice and soy), a positive yield effect is found. As both crop production and climate change are location specific, it is important to take spatial effects into account. Although maize is grown throughout all regions in Zambia, most is produced in the Eastern province for which yield loss is expected to be between 5 and 25%. In the Northern province, where there are potential yield gains, there is hardly any maize production because of the highly acidic soils (@Figure S1). For the other crops, yield change is more evenly distributed over the country which means location of production is less relevant.

_Figure 1: Yield change under climate change: 2000-2050_

```{r fig_yecc_ag}
fig_yecc_ag
```

##### Source: Crop model simulations
##### Note: Average of 2 crop models and 5 GCMs, using area harvested in 2000 as weights. Data are for subsistence farming systems.


### CSA scenarios

After consultation with stakeholders, we selected seven CSA options for the impact analysis. For each of the options we defined impact coefficents for crop yield and GHG using information from the literature and input from stakeholders [@SI contains additional information]. Table @2 summarizes the seven CSA strategies and provides information on the coefficients used  in GLOBIOM. 

[@ADD some description, explain that CA is broken down into three parts, 


The final impact of adopting a CSA option is determined by the combination of three factors: (1) The crop yield and GHG technical coefficients, which reflect the direct impact of the CSA strategy; (2) The adoption rate, which reflects the size of the area on which the CSA strategy is used and (3) The number of crops on which the CSA strategy is applied.  


_Table 1: Summary of CSA strategies_

```{r table_sum_csa}
library(readxl)
table_sum_csa <- read_excel(file.path(projectPath, "report/paper/Fig_tab_map/external/tables.xlsx"), sheet = "table_sum_csa")
pander(table_sum_csa, missing = "")
```

##### Source: FAO CSIP background report, @Thierfelder2013, @Frank2018, @Shitumbanuma2012, @Haggblade2003.


# Results

This section presents the results of our modelling effort. The first part of this section briefly describes the set of indicators we use to assess future agricultural development of Zambia. The second part shows the model projections for the BAU scenario up to 2050 and compares those with the stakeholder vision. Finally, we illustrate the impact of adopting the selected CSA technologies on key agro-economic indicators.

## Business as usual projections

The following section presents projections for a number of key agricultural variables up to the year 2050. The figures show the changes over time in terms of socioeconomic and agricultural development for the coming three decades assuming a Business as Usual (BAU) scenario for Zambia. The BAU represents a scenario of zero climate change. To show the impact of climate change, we added error bars to the figures, which show the range of projections under different climate change scenarios. As indicated by @FigureX, and @SI, for some regions and a selection fo GCMs, climate change can also have postive effects. As climate impact only starts to materialize in the long term, we only added error bars for the period 2020-2050. In the following section, we take a closer look at the impact of using various CSA technology options in comparison to the BAU scenario. This is followed by a section in which the results of a sensitivity analysis assessing the outcomes under different projections for population and economic growth. 

### Crop yields

Figure @25 depicts both the trend in historical crop yield from FAOSTAT and GLOBIOM projections for the four key food crops in Zambia: maize, cassava, groundnut and millet. GLOBIOM considers the technological improvement in crop productivity (@Figure 5) as a starting point for potential yield growth. Additionally, yields will reflect the change in cropland area between production systems and also the reallocation of crop production to more or less productive areas. Increases in production over time will come from a combination of the assumed yield improvements, expansion of the crop to current cropland including transitions between production systems, and expansion of the crop to new cropland taking into account the costs associate with land conversion and transitions between systems.

Yield projections produced by GLOBIOM for the period 2000-2011 are very similar to the historical trend from FAOSTAT. Only in case of cassava, the yield projections might be too optimistic. In the BAU, maize yields more than double from 2000 to 2050 reaching about 3.9 fm t/ha. This is comparable to results of @Kanyanga2013, who presents maize yield projections in the range of 3 fm t/ha for 2050. Cassava yields increase 55%, groundnut yields increase almost 90% and millet yields increase 51% from 2000 to 2050. The error bars illustrate the potential range in crop yield when running a range of different climate change scenarios and using a variety of crop models. The error bars in Figure 25 show that cassava yield is more likely to increase than decrease under climate change, while the reverse is true for the other three crops. In the most extreme scenario, maize yield is projected to be 3.3 fm t/y, which is 15% lower than the projected yield in the baseline of 3.9 fm t/ha. 

_Figure 2: BAU yield_

```{r fig_bau_yld}
fig_bau_yld
```

##### Source: FAOSTAT (historical), IIASA (projections) and Table 1 (vision)


### Crop production 

The production of crops in Zambia is driven by changes in crop demand (also including foreign demand through trade) and supply side factors, in particular expected changes in crop yield. In 2000, maize accounted for nearly half of the total production of crop products in Zambia followed by sugar cane, cassava, wheat and groundnuts. Under the BAU scenario, total crop production doubles by 2050 reaching around 6.5 Mt dm (Figure 26). By 2050, maize is still the most produced crop in the region (near sevenfold production increase). The GLOBIOM projections for 2000-2010 are in line with the historical trend, which shows gradual expansion in production of maize, sugar and several other crops. Only after 2010, production, in particular that of maize, started to increase rapidly, probably as a consequence of agricultural subsidy programs. As this implies a diversion of the BAU, which is unlikely to be maintained for the coming four decades, this fluctuation is not taken into account by the model.

_Figure 3: BAU crop production_

```{r fig_bau_crop_prod}
fig_bau_crop_prod
```

##### Source: FAOSTAT (historical) and IIASA (projections)
##### Note: Error bars indicate the range in total crop production as a consequence of climate change.

### Food availability

According to the FAO, calorie availability in Zambia has hovered at or below 2,000 kilo calories per capita per day since around 1985 and in 2010, the per capita calorie consumption was about 1900 (FAO 2017). In the BAU scenario (Figure 31) we assume an evolution of diets in line with FAO projections [@Alexandratos2012]. The FAO projects a continuation of the diet transitions for developing countries such as Zambia, where calorie availability increases over the time due to the increase in the GDP per capita and results in an increase in animal products. By 2050, calories are about 40% higher than in 2000 (an increase of more than 800 kilo calories per day). Calorie availability is a measure of the total final demand of households or food available for consumption which does not include retail waste but does include household waste. In terms of food availability, FAO considers 2500-3000 kcals/capita/day to be a target for developing countries. @Kanyanga2013, projected a range of 1600-2400 kcal/cap/day for a only a subset of crops and assumes slower economic growth. Using the same modelling approach, @IFPRI2017 has more recent projections for food availability showing an increase in Zambia of kcals to 2250 in 2030 and 2640 by 2050. Our projections are nearly in line with these new IFPRI projections.

_Figure 4: BAU Calorie availability_

```{r fig_bau_cal}
fig_bau_cal
```

##### Source: FAOSTAT (historical), IIASA (projections) and Table 1 (vision)



### Agriculture GHG emissions 

The greatest contributor to total GHG emissions in Zambia is the land use, land-use change, and forestry (LULUCF) sector, accounting for 86.6% of total emissions [@CIAT2017]. Agriculture and energy are responsible for 6% and and 6.5% of total emissions, respectively. Within LULUCF, 93% of the emissions are caused by the burning of biomass, while only 7% is due to the conversion of forest into cropland or any other type of land use conversion. GHG emissions in the agricultural sector or equally concentrated as the burning of savanna is responsible for 59% of total emissions, followed by emissions from enteric fermentation (12%), manure left on pasture (10%), cultivation of organic soils (10%), the use of synthetic fertilizers (5%) and other sectors (i.e. other use of manure, rice cultivation and crop residues).

@Figure shows the GHG emissions from the agricultural sector under the BAU scenario. In 2000, livestock accounted for more than 97% of the total GHG emissions from agriculture (not taking into account emissions from cultivation of organic soils and crop residue), while the remainder was caused by the crop sector, in particular the use of synthetic fertilizers (2.5%) and the production of rice (0.5%). In 2050 the share of emissions from synthetic fertilizer has increased to over 13% of total emissions because of a shift toward more intensive production systems that make use of chemical fertilizers. The historical trend shows a much larger increase in total GHG emissions than our projections. A closer look at the data shows that according to FAOSTAT, the growth in emissions from synthetic fertilizers is in fact much higher than the BAU projection. As mentioned above, the most likely reason for this observation is the recent and rapid growth in the application rate of fertilizers on maize, triggered by the national maize subsidy program.

_Figure 5: BAU agricultural GHG Emissions_

```{r fig_bau_ag_emis}
fig_bau_ag_emis
```

##### Source: FAOSTAT (historical), IIASA (projections) and Table 1 (vision)


## CSA impact scenarios

This section reviews the impact of adopting the various CSA strategies discussed above on a number of key agricultural variables. In the figures below we present the BAU as "none" and then each bar represents a scenario in which one of the CSA practices is adopted using the assumptions described in Section X. In the figures the adoption of agroforestry will appear as "af", conservation agriculture as "ca", residue retention as "rr", minimum soil disturbance as "msd", drought tolerant maize as "dtm", post-harvest losses as "phl" and the diversification scenario as "div". In most cases, the figures show _the difference (in %)_ with the BAU projected values as described in the previous section. The error bars represent the maximum and minimum impacts from climate change. In most cases results are only presented for the five largest crops in Zambia: maize, cassava, groundnut, millet and sugar cane. We also present the results for cotton because the FAO survey shows that farmers in Zambia are actively applying CSA practices to the production of cotton as well as soy because it is one of the target crops in the diversification scenario.


### Crop yields

In figure 35 we compare average crop yield in 2050 under the different CSA options with the BAU. Strategies to reduce post-harvest losses have a large and positive impact for all crops. If we only look at maize, phl has the largest yield impact (increase of 23% in comparison to the BAU), followed by ca (14%), af (12%), msd (9%) and rr (4%). In case of cotton, msd (21%), phl (12%) and ca (11%) have the largest impact, followed by ca (3%) and rr (1%). Adopting drought tolerant maize varieties has no impact in comparison to the BAU under _no climate change_ because these varieties are assumed to have the same yield as the conventional varieties that are used by farmers. However, the error bar on the dtm bar for maize shows that a yield gain of 8% is achieved under the climate change scenarios. All other options are also sensitive to changes in climate as shown by the error bars. In the diversification scenario, the yield of soy beans increases with 9%, while the effect on other crops is very small. 

The reduction of post-harvest losses has the potential to substantially increase crop yields, thereby making a large contribution to achieving the vision of doubling crop yield in 2050.  For example, in case of maize, the adoption of phl will increase yields from 3.9 t/ha under the BAU to 4.8 t/h, which is very close to the vision target of 4.9 to/ha.

_Figure 6: Yield change in comparison to the BAU_

```{r fig_opt_yld,  fig.width=8, fig.height=12}
#fig_opt_yld_abs
fig_opt_yld_dif
```


### Crop production

Figure 36 depicts the change in crop production for the six selected crops as well as total agriculture. Not surprisingly, the figures show a similar pattern as the change in crop yield. Looking at total agricultural production (i.e. the sum of all crops that are modeled in GLOBIOM), phl (13%) will contribute the most of all CSA options to increasing crop output, followed by div (11%), ca (7%), msd and af (both 4%) and rr (1%). The impact of phl is very large because it affects all crops simultaneously, whereas the other CSA options only apply to a handful of crops. Shifting to drought-tolerant maize varieties under a climate change scenario increases agricultural production marginally (<1%) because this strategy only affects maize. The effect of climate change on the other crops will still be negative, resulting in a very small total effect. In the div scenario, production of groundnut (20%) and soy (25%) increase dramatically in comparison to the BAU.

_Figure 7: Crop production change in comparison to the BAU_

```{r fig_opt_prod,  fig.width=8, fig.height=12}
#fig_opt_prod_abs
fig_opt_prod_dif
```

##### Source: GLOBIOM


### Food availability

Figure 38 shows the change in kcal/cap/day, a proxy of food availability, in comparison with the BAU in the year 2050. phl (7%), and ca (5%) have the highest impact on food availability. The total effect is smaller than the increase in agricultural production (Figure 38) because not all production is for food consumption. 


_Figure 8: Food availability change in comparison to the BAU_

```{r fig_opt_calo}
fig_opt_calo_dif
```

##### Source: GLOBIOM


### GHG emissions

The impact of adopting CSA practices on GHG emissions runs through two channels: (1) a direct effect because some of the CSA practices will lead a change in GHG emissions per ha. As shown in Table 2, we assume that ca and msd practices lead to lower CO2eq emissions per ha, while adoption of rr practices is expected to increase the emissions per ha. The direct effect will be attributed to the category that measures the GHG emissions. (2) an indirect effect. For example, GHG emissions might be lower (or higher) because the adoption of CSA practices might result in lower (higher) expansion of cropland and, hence, less (more) land use change. Figure 40 shows that the land use change effects are very small. Not surprisingly the results show that GHG emissions in categories other than from synthetic fertilizers (i.e. livestock related categories and land use change) are negligible. For this reason we only focus on the synthetic fertilizers category here. 

Figure 40 shows that expansion of msd and phl practices results in decrease  of emissions from fertilizer use with 11% and 5%, respectively in comparison to the BAU. The impact of phl is indirect, where a  free productivity benefit decreases the increases the production while at the same time reduces the cropland area for maize, which results in a decrease in nitrogen fertilizer demand by 8%. In contrast with the other options, the expansion of rr and the implementation of diversification policies will lead to an increase in GHG emissions with 39% and 7% respectively. This clearly illustrates that some CSA solutions entail trade offs between increasing productivity and decreasing emissions. 

If we only consider emissions from synthetic fertilizer use, the stand-alone adoption of msd and phl practices by farmers in Zambia will result in reduction of GHG emissions but this will not be sufficient to reach the INDC targets for reducing GHG emissions by 25-47% (for the synthetic fertilizer category). It is, however, important to note that GLOBIOM might underestimate the GHG mitigation impact for some of the CSA options, in particular ca and possibly msd and af. There is evidence that implementing these strategies might have a positive effect on soil carbon sequestration [@Smith2008; @Corsi2012], which is not captured in GLOBIOM.

_Figure 9: Emissions from synthetic fertilizers change in comparison to the BAU_

```{r fig_opt_emis}
fig_opt_ag_emis_dif
```

##### Source: GLOBIOM


# Discussion

The final impact of adopting a CSA option is determined by the combination of three factors: (1) the assumption on (higher) productivity of the CSA option in comparison to conventional agriculture; (2) the projected adoption rates for each of the CSA strategies and (3) the selection of crops on which the practice are applied. Of the CSA options, phl can be applied to all crops. Most other  practices can only be applied to maize farming and a selection of other crops. In the case a CSA option-crop combination is not possible, the impact will show up as zero in the figures.


In this section we aim to compare the potential impact of the various CSA practices by means of a simple ranking exercise. At the moment we consider only two variables that can easily be compared: (1) total GHG emissions from synthetic fertilizers and (2) agricultural production. As in the previous section, all variables are measured as % change in comparison to the BAU. To account for climate change effects we rank the average impact by CSA option over all climate change scenarios (i.e. the values indicated by the errorbars) instead of the BAU that assumes no climate change impact (indicated by the bars). In practice, both ranking approaches give identical results. 

We decided to use these two variables because they can be compared over the total agricultural sector, taking into account the impact over all crops; something which is not possible with an indicator like crop yield. Moreover, the two indicators capture two of the three components of CSA. Production is a related to the 'productivity' component of CSA, while GHG emissions capture the 'mitigation' component. We do not have a good indicator for the 'adoption' component. To compare the practices we simply ranked them from 1 to 7, where 1 indicates the highest 'climate smart' impact (in comparison to the BAU), i.e. the highest production or the most emissions reduced. The total ranking is the unweighted average of the two rankings

Figure 41 depicts the GHG, production and total ranking. There are clearly some trade-offs between the two indicators for some practices. The most obvious case is div, which scores high on production but very low on GHG emissions.  Overall phl and msd have the highest ranking when both indicators are considered. The main reason why these two options are top ranking is their wide applicability (to all crops for phl and eight crops for msd) and the expected high future adoption rates (50% for msd). Conservation agriculture and, to a lesser extent, agro-forestry have high yield coefficients but because of the limited crop coverage and the expected low future adoption rates, projected final impact is low. 


##### Source: Authors

# Conclusions

In this paper we presented a scenario analysis and model study to assess the impact of the adoption of CSA options on agricultural development under climate change in Zambia. The results revealed differentiated impact of the seven selected CSA practices on agricultural production, crop yield, land use change, trade, food consumption and GHG emissions in comparison to a business as usual scenario. Differences are caused by three factors: (1) impact on yield and GHG emissions after adoption, (2) applicability, i.e. number of crops to which it can be applied, and (3) the adoption rate in 2050. Overall, the results show that, absent of climate change, adoption of all CSA strategies but drought tolerant maize varieties resulted in an increase in food security and in most cases less expansion of cropland area. Results also show that the adoption of post-harvest loss practices, minimum soil disturbance (no tillage) and conservation agriculture could result in the highest reductions of GHG emissions, while at the same time having a positive impact on crop yield and production. Adoption of agro-forestry and drought tolerant maize varieties have positive but relative lower impact on crop production and GHG emissions. Finally, policies to promote diversification of agriculture and adoption of residue retention practices have a mixed effect, resulting in higher crop production accompanies by increases GHG emissions. Policies to promote CSA in Zambia should be aware about the differences and trade-offs in terms of climate mitigation and productivity impacts between CSA practices. 

When considering the results it is important to take the following limitations of the analysis into account. First, to model and compare the future impact of CSA practices, we need coefficients on how crop yield, production and GHG emissions are affected when these practices are adopted in comparison with conventional farming practices. Together with the assumptions on adoption rates, change in key drivers and market effects, these coefficients will determine how much impact the strategy will have on productivity, climate resilience and mitigation. Such information can only be derived from detailed farm and plot level studies that use a (semi-) experimental setting to systematically compare the impact of different CSA options, while controlling for external conditions. Surprisingly, we found only a few studies that present this information and in most cases they covered only one CSA strategy and/or crop, making it difficult to compare due to differences in methodologies and described study settings outside Zambia. Hence, the indicators we use are a proxy at best and therefore results should be interpreted with care. More farm and plot level studies are needed to assess  the impact of CSA practices on yield and GHG emissions are needed to improve the analysis.

Second, we are not able to fully take into account all spatially explicit factors that determine the impact of CSA strategies on agricultural production. As mentioned in the previous point, we have only very crude proxy indicators to differentiate the impact of CSA options in terms of crop yield and GHG emissions. In practice, the impact of adopting conservation agriculture, agro-forestry or drought tolerant maize varieties will be different across Zambia, depending among others on soil quality, climate and learning effects. Similarly, due to lack of high-resolution data on the location of smallholder farmers, spatial distribution of crops and costs of inputs, the model is run at a relatively coarse resolution (200 x 200 km), which limited spatial representation.

Third, the key source of GHG emissions in Zambia is the burning of biomass, which makes up around 80% of total GHG emissions in the country. Another important cause for these emissions is the burning of forests for the production of charcoal, which is an important source of household energy supply in Zambia [@Dlamini2016]. Burning of savanna is another source of GHG emissions (3.5%). At the moment GLOBIOM is not able to project the change in these two types of emission due to lack of information. The modelling of the burning of biomass would require projections on the demand for charcoal, which, in turn, depends on the future supply of household energy in Zambia (e.g. hydropower, solar and wind energy). An interesting avenue for future research would be the design and modelling of such scenarios.

Fourth, the crop model output we use to prepare our climate change crop yield shocks only capture the impact of anthropogenic climate change, i.e. the long-run changes in climate caused by human activity. This is different from climate variability (e.g. extreme weather events such as floods, droughts and hurricanes), which is also expected have considerable and negative impact on future economic and agricultural development in Sub-Saharan Africa [@Ahmed2011; @Thurlow2012]. Modelling the effects of year-by-year variations in climate requires the use of a stochastic modelling approach, which is currently under development.

Finally, we were unable to find detailed comparable information on the costs of adopting the CSA options and therefore imposed an exogenous rate of adoption to model their diffusion. Despite a long tradition, the present adoption rate of conservation agriculture remain low in Zambia [@Zulu-mbata2016]. Main reasons for the limited uptake include credit constraints, social and cultural beliefs, labor costs and competition for crop residues [@Arslan2013;@Zulu-mbata2016]. This illustrates that the adoption of CSA is not easy nor automatic. Policies that aim to promote the spread of CSA practices in Zambia should take this into account and include measures to facilitate adoption, such as the provision of extension services, training and credit provision.     


# References


